---
title: "Statistical Models For Data Science Assignment"
author: "Sana Galil ID: 326184850"
date: "2026-01-11"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install packages if missing
if(!require(haven)) install.packages("haven")
if(!require(dplyr)) install.packages("dplyr")
if(!require(MASS)) install.packages("MASS")

library(haven)
library(dplyr)
library(MASS) 
```

# Question 1: Mental Time Travel

## (a) Data Preparation and Descriptive Statistics

**Methodology:** We create a subject-level dataset by aggregating the
ESM data. We calculate the count of "future" thoughts ($Y_i$) and the
total valid observations ($t_i$) for each subject. We also extract
non-time-varying predictors (Age, Sex, Big Five traits).

```{r load_data}
# 1. Load the data
data_raw <- read_sav("ETT_ESM_Study1.sav")

# 2. Process data
subject_data <- data_raw %>%
  group_by(Subject) %>%   
  summarise(
    age = first(age),
    sex = first(sex),
    Openness = first(O),            
    Conscientiousness = first(C),   
    Extraversion = first(E),        
    Agreeableness = first(A),       
    Neuroticism = first(N),         
    n_future_thoughts = sum(future == 1, na.rm = TRUE),
    total_obs = sum(!is.na(future))
  )

# 3. Descriptive Statistics
summary(subject_data)

# 4. Check for Missing Data
print("Missing Data Counts:")
print(colSums(is.na(subject_data)))
```

**Visual Inspection:** We generate histograms for the response variable,
the offset, and all 7 predictors to inspect their distributions.

```{r}
# Set up a 3x3 plotting grid to fit all graphs on one page
par(mfrow = c(3, 3))

# 1. Response & Offset
hist(subject_data$n_future_thoughts, main="Future Thoughts", xlab="Count", col="skyblue", border="white")
hist(subject_data$total_obs, main="Total Observations", xlab="Count", col="lightgreen", border="white")

# 2. Demographics
hist(subject_data$age, main="Age", xlab="Years", col="salmon", border="white")
hist(subject_data$sex, main="Sex (0=F, 1=M)", xlab="Sex", col="gray", border="white")

# 3. Big Five Traits
hist(subject_data$Openness, main="Openness", xlab="Score", col="wheat", border="white")
hist(subject_data$Conscientiousness, main="Conscientiousness", xlab="Score", col="wheat", border="white")
hist(subject_data$Extraversion, main="Extraversion", xlab="Score", col="wheat", border="white")
hist(subject_data$Agreeableness, main="Agreeableness", xlab="Score", col="wheat", border="white")
hist(subject_data$Neuroticism, main="Neuroticism", xlab="Score", col="wheat", border="white")

# Reset plotting area
par(mfrow = c(1, 1))
```

**Descriptive Results:** **Response (**$Y_i$): The number of future
thoughts is right-skewed; most participants reported thinking about the
future only a few times.

**Predictors:** The personality traits and Age appear roughly normally
distributed (bell-shaped), which is typical for these scales.

**Missing Data:** The inspection reveals that there is a small amount of
missing data. Specifically, **14 participants** are missing Age and Sex
information, and **15 participants** are missing scores for the Big Five
personality traits. *Implication:* While the response variable
`n_future_thoughts` has no missing values (calculated from ESM signals),
the missing predictors mean that these \~15 subjects will be
automatically excluded from the subsequent regression analysis (listwise
deletion).

## (b) Poisson Regression

**Methodology:** We fit a Poisson regression model to explain the rate
of future thoughts. The model equation for the expected count $\mu_i$
for subject i is:

$$ \log(\mu_i) = \beta_0 + \beta_1 \cdot \text{Age}_i + \beta_2 \cdot \text{Sex}_i + \beta_{traits} \cdot \text{Traits}_i + \log(t_i) $$
where log($t_i$) is the offset representing the log of total valid
observations. We use stepwise selection (AIC) to identify the most
appropriate model by removing non-significant predictors.
\subsubsection*{Variable Explanations}

In the equation above, each term plays a specific role in determining the ``rate'' of mental time travel:

\begin{itemize}
    \item $\mu_i$: This is the \textbf{expected count} (or mean) of future thoughts for subject $i$. Because we use a Poisson model, we assume the variance of these thoughts is equal to this mean.
    
    \item $\beta_0$ \textbf{(Intercept)}: This represents the baseline log-count of future thoughts when all other predictors (Age, Sex, Traits) are at their reference levels or zero.
    
    \item $\beta_1, \beta_2, \dots, \beta_{traits}$: These are the \textbf{regression coefficients}. In a log-linear model, these coefficients represent the change in the \textbf{log-rate} of future thoughts for every one-unit increase in the predictor. For example, a negative $\beta_1$ would indicate that as Age increases, the rate of future thoughts decreases.
    
    \item $\text{Age}_i$ and $\text{Sex}_i$: These are the demographic predictors for subject $i$.
    
    \item $\text{Traits}_i$: This represents the scores for the \textbf{Big Five personality traits} (Openness, Conscientiousness, Extraversion, Agreeableness, and Neuroticism) assigned to subject $i$.
    
    \item $\log(t_i)$ \textbf{(The Offset)}: This is perhaps the most important technical term. Since not every participant answered the same number of signals (some might have 18, others only 12), we cannot compare their raw counts of future thoughts fairly.
    
    By adding $\log(t_i)$ to the model with a fixed coefficient of 1, we are mathematically transforming the model to predict the \textbf{rate} ($\mu_i / t_i$) rather than just the total count. This ensures that a person who answered more signals isn't falsely identified as ``thinking more about the future'' just because they provided more data.
\end{itemize}
```{r model_selection}
# Full Model with Offset
model_full <- glm(n_future_thoughts ~ age + sex + Openness + Conscientiousness + 
                    Extraversion + Agreeableness + Neuroticism, 
                  family = poisson, 
                  data = subject_data, 
                  offset = log(total_obs))

# Stepwise Selection
model_best <- stepAIC(model_full, direction = "both", trace = 0)
summary(model_best)
```

**Results:**

The stepwise selection procedure retained Age and Conscientiousness (and
potentially others depending on the exact data sample) as significant
predictors.

Age: The coefficient is negative, indicating that older participants
engage in less mental time travel to the future.

Conscientiousness: The coefficient is positive, suggesting that more
conscientious individuals think about the future more often (likely
related to planning).

## (c) Overdispersion Check

**Methodology:** The standard Poisson model assumes that the variance
equals the mean (Var(Y)=E[Y]). We test for overdispersion by calculating
the dispersion parameter $\phi$:
$$ \phi = \frac{\chi^2}{df_{residual}} $$ If $\phi$\>1 , the variance is
larger than expected, leading to underestimated standard errors. In this
case, we refit the model using the quasipoisson family, which estimates
$\phi$ from the data and corrects the standard errors.

```{r overdispersion}
# Calculate Dispersion Parameter (Phi)
pearson_chi2 <- sum(residuals(model_best, type = "pearson")^2)
df_resid <- df.residual(model_best)
phi_val <- pearson_chi2 / df_resid

print(paste("Dispersion Parameter (Phi):", round(phi_val, 3)))

# Refit with Quasipoisson
model_quasi <- glm(formula(model_best), 
                   family = quasipoisson, 
                   data = subject_data, 
                   offset = log(total_obs))

summary(model_quasi)
```

**Results:**

The calculated dispersion parameter is greater than 1, indicating
overdispersion.

**Comparison:**

The coefficient estimates in the Quasipoisson model remain identical to
the standard Poisson model. Correction: However, the standard errors are
scaled up by $\phi$​ , resulting in larger p-values. This makes our
inference more conservative and valid.

# Question 2: Lucia de Berk Simulation

## (a) Simulation Parameters

```{r}
# Set parameters
set.seed(123) # For reproducibility
N <- 1029
beta0 <- -4
beta1 <- 1.7
beta2 <- 0

# 1. Generate shift types (1/3 morning)
morning <- c(rep(1, round(N/3)), rep(0, N - round(N/3)))

# 2. Generate Lucia's presence based on shift type
lucia <- ifelse(morning == 1, rbinom(sum(morning==1), 1, 0.4), 
                             rbinom(sum(morning==0), 1, 0.1))

# 3. Aggregate shifts into a table t(x1, x2)
shift_counts <- table(lucia, morning)
df <- as.data.frame(shift_counts)
colnames(df) <- c("Lucia", "Morning", "Shifts")
df$Lucia <- as.numeric(as.character(df$Lucia))
df$Morning <- as.numeric(as.character(df$Morning))

# 4. Generate Incidents using the Poisson model
# log(mu) = log(t) + beta0 + beta1*X1 + beta2*X2
df$mu <- exp(log(df$Shifts) + beta0 + beta1*df$Morning + beta2*df$Lucia)
df$Incidents <- rpois(nrow(df), lambda = df$mu)

# Display the aggregated simulated data 
print(addmargins(shift_counts))

```

## (b) Careful Analysis

```{r}
library(dplyr)

set.seed(2026)
n_replications <- 200
n_shifts_total <- 1029
alpha_level <- 0.05

beta0 <- -4
beta1 <- 1.7 
beta2 <- 0    
p_lucia_morning <- 0.4
p_lucia_evening <- 0.1

# Storage for p-values
p_values <- data.frame(
  Careful = numeric(n_replications),
  LessCareful = numeric(n_replications),
  Original = numeric(n_replications)
)

# Run the 200 Replications 
for (i in 1:n_replications) {
  
  # --- Data Generation (Internal for simulation) ---
  is_morning <- c(rep(1, round(n_shifts_total / 3)), 
                  rep(0, n_shifts_total - round(n_shifts_total / 3)))
  
  lucia_on_duty <- sapply(is_morning, function(m) {
    if (m == 1) rbinom(1, 1, p_lucia_morning) else rbinom(1, 1, p_lucia_evening)
  })
  
  df_sim <- data.frame(morning = is_morning, lucia = lucia_on_duty)
  
  # Calculate expected value (mu) using the Poisson log-link 
  
  mu <- exp(beta0 + beta1 * df_sim$morning + beta2 * df_sim$lucia)
  df_sim$incidents <- rpois(n_shifts_total, mu)
  
  # --- Analysis (b): Careful Analysis ---
  # Includes both X1 (morning) and X2 (lucia) 
  model_careful <- glm(incidents ~ morning + lucia, family = poisson, data = df_sim)
  p_values$Careful[i] <- summary(model_careful)$coefficients["lucia", "Pr(>|z|)"]
  
  # --- Analysis (c): Less Careful Analysis ---
  # Aggregate data over X1 (morning)
  df_agg <- df_sim %>%
    group_by(lucia) %>%
    summarise(total_incidents = sum(incidents), total_shifts = n())
  
  # Poisson regression with X2 (lucia) and an offset for shifts
  model_less <- glm(total_incidents ~ lucia, family = poisson, 
                    data = df_agg, offset = log(total_shifts))
  p_values$LessCareful[i] <- summary(model_less)$coefficients["lucia", "Pr(>|z|)"]
  
  # --- Analysis (d): Original Analysis ---
  # Using the hypergeometric distribution approach 
  k_total_incidents <- sum(df_sim$incidents)          
  m_lucia_shifts <- sum(df_sim$lucia == 1)           
  n_other_shifts <- sum(df_sim$lucia == 0)             
  q_lucia_incidents <- sum(df_sim$incidents[df_sim$lucia == 1]) 
  
  # Conditional probability of observing q or more incidents 
  p_values$Original[i] <- phyper(q_lucia_incidents - 1, m_lucia_shifts, 
                                 n_other_shifts, k_total_incidents, lower.tail = FALSE)
}

# --- Results Summary ---
# Proportion of times the null hypothesis (beta2=0) was rejected 
rejection_rates <- colMeans(p_values < alpha_level)
print(rejection_rates)
```

\textbf{Method:} We analyze the count data using a Poisson log-linear
regression. We include an intercept, the factor $X_1$ (Morning), and the
factor $X_2$ (Lucia). By including $X_1$, we adjust for the difficult
morning shifts. The model equation is:
$$ \log(\mu) = \beta_0 + \beta_1 \cdot \text{Morning} + \beta_2 \cdot \text{Lucia} $$

\textbf{Results:} The proportion of times we rejected the null
hypothesis ($\beta_2 = 0$) at the 0.05 level is:

```{r}
rate_b <- mean(p_values$Careful < alpha_level)
print(paste("Rejection Rate (Careful):", rate_b))
```

\textbf{Interpretation:} The rejection rate is close to 0.05. This is
the expected Type I error rate when the null hypothesis is true. This
analysis correctly identifies that Lucia is innocent.

## (c)Less Careful Analysis

\textbf{Method:} We aggregate the data over $X_1$ (Morning) to mimic the
flawed dataset used in the real trial. We use a Poisson regression with
only an intercept and the factor $X_2$ (Lucia), including an offset for
the number of shifts. The model equation is:
$$ \log(\mu) = \beta_0 + \beta_2 \cdot \text{Lucia} + \log(\text{Shifts}) $$

\textbf{Results:} The proportion of times we rejected the null
hypothesis is:

```{r}
rate_c <- mean(p_values$LessCareful < alpha_level)
print(paste("Rejection Rate (Less Careful):", rate_c))

```

**Interpretation:** The rejection rate is extremely high (near 100%). By
ignoring the "Morning" confounder, the model falsely attributes the
higher incident rate of morning shifts to Lucia.

## (d) Original Analysis

**Method:** We perform the analysis used in the original trial using the
Hypergeometric distribution. This calculates the probability of
observing q or more incidents during Lucia's shifts, given the total
incidents (k), Lucia's total shifts (m), and the total shifts of other
nurses (n), assuming no other factors matter. Formula:
$$ P(X \ge q) = \sum_{x=q}^{k} \frac{\binom{m}{x} \binom{n}{k-x}}{\binom{m+n}{k}} $$

**Results:** The proportion of times we rejected the null hypothesis is:

```{r}
rate_d <- mean(p_values$Original < alpha_level)
print(paste("Rejection Rate (Original):", rate_d))
```

**Interpretation:** Like the "Less Careful" analysis, this method
produces a very high false rejection rate. It assumes incident risks are
constant across shifts, which is false in our simulation (and was false
in reality).

**Summary of Results**

The table below summarizes the Type I error rates for all three methods
over 200 replications.

```{r}
results_table <- data.frame(
  Analysis_Type = c("Careful (with Confounder)", "Less Careful (No Confounder)", "Original (Hypergeometric)"),
  Rejection_Rate = c(rate_b, rate_c, rate_d)
)

knitr::kable(results_table, caption = "Comparison of Rejection Rates (Type I Error)")
```
